Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-10 _paper.rmd
-updates to use cases, videos, captions, and text. TODO: regression Use cases (not faceting correctly) & videos.

(ref:casepenguins) Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

## 2) Chocolates, milk/dark chocolate classification

The chocolates dataset consists of 88 observations of 10 nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. We can see if a manufacturer accurately portrays the chocolate with this data. We are curious to see if chocolates that nutritionally look like milk chocolates are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We selected a chocolate labeled dark, through predicted to be milk chocolate compared with a chocolate labeled 85% cocoa.

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate in light of the values on the nutritional label. We vary the contribution of calories from fat. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points. In this case, it feels strange to call the selected observation a misclassification of the model. There are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. This feels more like a measurement theory-related problem. In this case, the candy is being sold as dark chocolate, while nutritional value more closely resembles milk chocolates.

## 3) FIFA, wage regression

@@

(ref:casefifa) FIFA 2020 data, a random forest model regresses wages [2020 Euros] from aggregated of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). The animate radial tour can be found at [shorturl.at/lCK26](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, offense skills are almost singularly important for explaining the offensive player. We vary the contribution of offensive skills. Offensive skills are removed in the tour (panel b, frame 3), and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

## 4) Ames housing 2018, sales price regression

@@

(ref:caseames) Ames housing 2018 regressing sales price [2018 USD]. Because the SHAP values are relatively well distributed across the variables, there is a lot of redundant information to explain the gap between select observations. This case may not be ideal for the cheem methodology, though we did find a near singular frame in the manual tour not, something we came looking for, but certainly, an exciting feature to find. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

>>>>

(ref:casepenguins) Species classification of Palmer penguin data. We select a adelie penguin that is mislabeled as an chinstrap. Starting from the attribution projection we varying the contribution to `bill length`. This explanation is not supported sense by the time there is moderate contribution. The current basis is fairly close to the regularly attribution to other chinstrap penguins aside from the sign on flipper length being flipped. It feels odd that SHAP has identified a profile where chinstrap and adelie penguins are hard to distinguish when there are separating orientations that should be found within 125 trees. The .mp4 animation of this tour can be found at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

## Chocolates, milk/dark chocolate classification

The chocolates dataset consists of 88 observations of ten nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. We can see if a manufacturer accurately portrays the chocolate with this data. We are curious to see if chocolates that nutritionally look like milk chocolates are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We selected a chocolate labeled dark, through predicted to be milk chocolate compared with a chocolate labeled 85% cocoa.

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove a 4 variables with the lowest contribution for the selected instance and vary the contribution of `Sugar`. The misclassification of two instances seems improbable when sugar is near max contribution. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `sugar`. We see notice 2 instances that have clearly well within the milk chocolate dispite being observed as dark chocolate on their product label. While conjector, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. This feels related a measurement theory problem.

## FIFA, wage regression

@@

(ref:casefifa) FIFA 2020 data, a random forest model regresses wages [2020 Euros] from nine aggregated of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three variables with low attribution from both players. The attribution projection starts with selected instance on the right. We vary the contribution from `defense`, the star offensive player is not distinguished in the horizontal direction. At this point defensive players have been rotated to the highest horizonatal value. The animate radial tour can be found at [shorturl.at/lCK26](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. While `offense` and `reaction` skills both crutial to the explaintion of a star offensive player. We increase the contribution of denfensive skills and Messi is no longer separated from the group while van Dijk and other defensive players are better explained in this case.

## Ames housing 2018, sales price regression

@@

(ref:caseames) Ames housing 2018 regressing sales price [2018 USD]. The primary observation is a house that a large `lot area` (`LtA`), but small `living area` (`LvA`). The comparison instance is in the opposite; a large house on a relatively small lot. Similar to the soccer players, these house sale have diverging attribution to better predict their sales prices. Varying the contribution `living area` the separation between these house sales crosses when there is a low contribution which is so important to explaining their prediction though in differet ways. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```
